Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-08 submission/manuscript.Rmd
Add female section unedited

@@

>>>>

## Females have increased sensitivity to germline gene misexpression with _Kdm5c_ loss
```{r, DESeq2_params_XX, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
#get the parameters used repeatedly, like the log2fold change cut off
source("../../KDM5C_Germ_Dev_Sex_RA/code/utilities/parameters.R")
```

- Sex influences the degree of Kdm5c-KO germline gene misexpression but not the sex of germ cell-enriched genes

It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repression have been conducted exclusively in males. Sex is particularly pertinent in the case of KDM5C, as KDM5C is more highly expressed in females than males. We assessed germline gene misexpression in _Kdm5c_ mutant male and female mouse epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo that reflects when germline genes are typically silenced in somatic cells.

We first identified differentially expressed genes (DEGs) compared to wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs and then filtered for germline-enriched genes using our previously curated list. While both male and female _Kdm5c_ mutants misexpressed germline genes, about twice as many were dysregulated in homozygous and heterozygous females compared to homozygous males (XX KO: , XX HET: , XY KO: ) (Figure 1A).

A significant portion of germline genes were dysreguated in both male and female _Kdm5c_ mutants (Figure 1B). We then compared the log 2 fold change of shared germline DEGs to assess if sex impacts the degree of germline gene misexpression. 
	* Results of this, are females more sensitive? (Figure 1C)

We then wondered if the increased number of germline DEGs in females could be caused by improper X chromosome inactivation (XCI), as the X chromosome is enriched for many testis-specific germline genes. However, both shared and female-specific germline DEGs did not display a significant bias towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 1E). Altogether these results indicate female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, likely independent of potential XCI defects.

## Germline gene misexpression in _Kdm5c_ mutants is independent of germ cell sex

While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. Many of the germline genes characterized in the male _Kdm5c_-KO brain have sperm-specific functions. It is currently unclear if sperm genes are also dysregulated in female _Kdm5c_ mutants, or if they would instead misexpress egg-biased genes.

To comprehensively assess the impact of gonadal sex on germline gene misexpression, we filtered our list of germline-enriched genes for egg (XX) or sperm (XY) biased genes. We determined male and female biased expression  using RNAseq data from wild-type male and female mice at embryonic day 12, 14, and 16. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased 
germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017] (Figure 2A).

@@

\begin{figure}
  \centering 
  \includegraphics{../../KDM5C_Germ_Dev_Sex_RA/submission/compiled_figs/EpiLC_XXvsXY.pdf}
  \caption[Sex influences the degree of Kdm5c-KO germline gene misexpression but not the sex of germ cell-enriched genes]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B). Left - MA plot of mRNA-sequencing. Right - Number of tissue-enriched differentially expressed genes (DEGs). * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal). Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red. \textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}. Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red. \textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}. Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
  \label{figurelabel}
\end{figure}
